### 2020-04-18 serial.tex
really need to go to bed

\begin{figure}[!th]
\includegraphics[width=\textwidth]{forward.pdf}
\caption{
\textbf{Epidemiological dynamics and changes in mean forward and backward delay distributions.}
A. Cumulative incidence over time.
B-D. Changes in the mean forward incubation period, generation interval, and serial interval.
E-G. Changes in the mean backward incubation period, generation interval, and serial interval.
Black lines represent the results of a deterministic simulation.
Orange lines and points represent the average of 10 stochastic simulations;
cumulative incidence from the stochastic simulations in panel A are shown without averaging.
Forward incubation periods and intrinsic generation-intervals are assumed to be independent of each other: $h(x_0, \sigma) = k(x_0) g(\sigma)$.
See Table 1 for parameter values.
}
\label{fig:epi}
\end{figure}

@@

\begin{figure}[!th]
\includegraphics[width=\textwidth]{observedrR.pdf}
\caption{
\textbf{Estimates of the reproduction number from the observed serial intervals.}
}
\label{fig:obsrR}
\end{figure}

Now, we turn to practical issues.
When an epidemic is ongoing, the observed serial intervals are subject to right-censoring because we cannot observe a serial interval if either an infector or an infectee has not yet developed symptoms.
\fref{obsrR} demonstrates how the effect of right-censoring in the observed serial intervals translates to the underestimation of $\mathcal R$.
Notably, even if we can observe \emph{all} serial intervals across all transmission pairs after the epidemic has ended, we still underestimate $\mathcal R$ by a large amount because the observed serial-interval distribution does not account for changes in the forward serial-interval distribution.
As the mean forward serial-interval distribution decreases over time, taking the average of all observed serial intervals throughout an epidemic will underestimate the mean forward serial-interval distribution during the exponential growth phase, which provides the correct linke between $r$ and $\mathcal R$.

We provide a simple, heuristic way of assessing potential biases in the estimate of $\mathcal R$ retrospctively.
Once a set of serial intervals has been observed, we can group observed serial intervals by their primary cohort times.
Then, we can compare how estimates of $\mathcal R$ change as we include more cohorts into the analysis (see `cohort-averaged' in \fref{obsrR}).
During the exponential growth phase, the estimates of $\mathcal R$ are consistent with the true value;
adding more data allows us to make more precise inference about $\mathcal R$ during this period.
However, the cohort-averaged estimates of $\mathcal R$ decreases rapidly soon after the exponential growth period.
Confidence intervals associated with the cohort-averaged estimates of $\mathcal R$ becomes narrower and do not overlap with earlier estimates.
This approach allows us to detect how changes in the forward serial intervals bias the estimates of $\mathcal R$.

>>>>

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=0.95\textwidth]{forward.pdf}
\caption{
\textbf{Epidemiological dynamics and changes in mean forward and backward delay distributions.}
A. Cumulative incidence over time.
B-D. Changes in the mean forward incubation period, generation interval, and serial interval.
E-G. Changes in the mean backward incubation period, generation interval, and serial interval.
Black lines represent the results of a deterministic simulation.
Orange lines and points represent the average of 10 stochastic simulations;
cumulative incidence from the stochastic simulations in panel A are shown without averaging.
Forward incubation periods and intrinsic generation-intervals are assumed to be independent of each other: $h(x_0, \sigma) = k(x_0) g(\sigma)$.
See Table 1 for parameter values.
}
\end{center}
\label{fig:epi}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{observedrR.pdf}
\caption{
\textbf{Estimates of the reproduction number from the observed serial intervals.}
A. Changes in the observed and cohort-averaged mean serial interval over time.
B. Changes in the estimate of $\mathcal R$ based on the observed and cohort-averaged mean serial interval over time.
Each line represents an independent stochastic realization.
}
\label{fig:obsrR}
\end{figure}

Now, we turn to practical issues.
When an epidemic is ongoing, the observed serial intervals are subject to right-censoring because we cannot observe a serial interval if either an infector or an infectee has not yet developed symptoms.
\fref{obsrR} demonstrates how the effect of right-censoring in the observed serial intervals translates to the underestimation of $\mathcal R$.
Notably, even if we can observe \emph{all} serial intervals across all transmission pairs after the epidemic has ended, we still underestimate the initial mean forward serial interval (\fref{obsrR}A) and therefore $\mathcal R_0$ (\fref{obsrR}) by a large amount because the observed serial-interval distribution does not account for changes in the forward serial-interval distribution.
As the mean forward serial-interval distribution decreases over time, taking the average of all observed serial intervals throughout an epidemic will underestimate the mean of the initial forward serial-interval distribution, which provides the correct linke between $r$ and $\mathcal R$.

We provide a simple, heuristic way of assessing potential biases in the estimate of $\mathcal R$ \emph{retrospctively}.
Once serial intervals have been observed after the epidemic has been sufficiently progressed, we can group observed serial intervals by their primary cohort times.
Then, we can compare how estimates of the mean serial interval and $\mathcal R$ change as we include more cohorts into the analysis (see `cohort-averaged' in \fref{obsrR}).
During the exponential growth phase, the estimates of the mean serial interval and $\mathcal R$ are consistent with the target value;
adding more data allows us to make more precise inference during this period.
However, the cohort-averaged estimates decreases rapidly soon after the exponential growth period.
This approach allows us to detect how changes in the forward serial intervals bias the estimates of $\mathcal R$.
